---
title: "SNU601 gastric carcinoma — scatools depth CNV"
subtitle: "Highly aneuploid tumor benchmark (contrast to the flat PBMC normal)"
author: "scatools benchmark"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    embed-resources: true
    fig-width: 10
    fig-height: 5
execute:
  warning: false
  message: false
  cache: false
---

## Overview

Runs the `scatools` **depth-based** copy-number pipeline on real SNU601 gastric
carcinoma scATAC-seq — the shared benchmark of epiAneufinder and Alleloscope,
with matched single-cell DNA truth (Andor 2020). Unlike the PBMC negative
control (which should be ~flat), SNU601 is highly aneuploid, so this exercises
the method on genuine CNV signal.

Input is a pre-binned **200 kb bin × cell** depth matrix from the Alleloscope
repository (`data-raw/SNU601/scATAC/chr200k_fragments_sub.txt`), so we bypass
`bin_atac_frags()` and enter the pipeline at the SingleCellExperiment stage.

```{r setup}
library(scatools)
library(SingleCellExperiment)
library(Matrix)
library(BiocParallel)

bp <- MulticoreParam(workers = 8)

root <- file.path(
  Sys.getenv("SCATOOLS_BENCH_DATA", "~/work/scatools_benchmark_data"),
  "snu601", "Alleloscope", "data-raw", "SNU601", "scATAC"
)
stopifnot(dir.exists(root))
```

## Load the pre-binned tumor matrix

```{r load}
barcodes <- readLines(file.path(root, "barcodes.tsv"))

dt <- data.table::fread(file.path(root, "chr200k_fragments_sub.txt"),
  header = FALSE, skip = 1
)
bin_ids <- dt[[1]]
counts <- as(as.matrix(dt[, -1, with = FALSE]), "CsparseMatrix")
colnames(counts) <- barcodes
rownames(counts) <- bin_ids

# Parse "chr1-218200001-218400000" -> GRanges
parts <- do.call(rbind, strsplit(bin_ids, "-", fixed = TRUE))
gr <- GenomicRanges::GRanges(
  seqnames = parts[, 1],
  ranges = IRanges::IRanges(start = as.numeric(parts[, 2]), end = as.numeric(parts[, 3]))
)
names(gr) <- bin_ids

# Keep standard chromosomes and clip bins to chromosome lengths (the terminal
# 200 kb bin of each chromosome overshoots the sequence boundary).
bsg <- BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38
keep <- as.character(GenomicRanges::seqnames(gr)) %in% GenomeInfoDb::seqnames(bsg)
gr <- gr[keep]
counts <- counts[keep, , drop = FALSE]
GenomeInfoDb::seqlevels(gr) <- GenomeInfoDb::seqlevelsInUse(gr)
GenomeInfoDb::seqlengths(gr) <- GenomeInfoDb::seqlengths(bsg)[GenomeInfoDb::seqlevels(gr)]
gr <- GenomicRanges::trim(gr)

dim(counts)
```

## Per-bin GC content (hg38)

```{r gc}
gseq <- BSgenome::getSeq(bsg, gr)
freqs <- Biostrings::alphabetFrequency(gseq)
gr$gc <- (freqs[, "C"] + freqs[, "G"]) / rowSums(freqs)
gr$n_freq <- freqs[, "N"] / rowSums(freqs)
gr$bin_id <- bin_ids

sce <- SingleCellExperiment(assays = list(counts = counts), rowRanges = gr)
sce <- sort(sce)
sce
```

## scatools depth-CNV pipeline

Bins are uniform 200 kb, so length normalization is a no-op and is skipped.

```{r pipeline}
sce <- filter_sce(sce,
  gc_range = c(0.3, 0.8),
  min_bin_counts = 1, min_bin_prop = 0.9,
  min_cell_counts = 1, min_cell_prop = 0.9
)
sce <- scatools:::add_ideal_mat(sce, ncores = 8)
sce <- add_gc_cor(sce, method = "modal", bpparam = bp)
sce <- smooth_counts(sce, assay_name = "counts_gc_modal", ncores = 8)
sce <- scatools:::calc_ratios(sce, assay_name = "counts_gc_modal_smoothed")
sce <- logNorm(sce, assay_name = "counts_gc_modal_smoothed_ratios", name = "logr_modal")
sce <- cluster_sce(sce, assay_name = "counts_gc_modal_smoothed_ratios", resolution = 0.3, verbose = FALSE)

# Segmentation + merge + normal-cell identification
sce <- segment_cnv(sce, assay_name = "counts_gc_modal_smoothed", bpparam = bp)
sce <- merge_segments(sce,
  smooth_assay = "counts_gc_modal_smoothed",
  segment_assay = "counts_gc_modal_smoothed_segment", bpparam = bp
)
sce <- identify_normal(sce, assay_name = "segment_merged_logratios", group_by = "clusters", method = "gmm", plot = FALSE)

cat("cells:", ncol(sce), "| bins:", nrow(sce), "| clusters:", length(unique(sce$clusters)), "\n")
```

## Aneuploidy signal

For a highly aneuploid tumor we expect strong, chromosome-scale gains/losses —
the opposite of the flat PBMC normal.

```{r signal}
lr <- as.matrix(assay(sce, "segment_merged_logratios"))
cat("median per-cell mean|logr|:", round(median(colMeans(abs(lr), na.rm = TRUE)), 3), "\n")
cat("fraction of bins with |mean logr| > 0.3:",
  round(mean(abs(rowMeans(lr, na.rm = TRUE)) > 0.3, na.rm = TRUE), 3), "\n")
```

## Copy-number heatmap

```{r heatmap, fig.height = 7}
sce <- sce[, order(sce$clusters)]
cnaHeatmap(sce,
  assay_name = "segment_merged_logratios",
  col_fun = logr_col_fun(),
  use_raster = FALSE
)
```

## Session info

```{r session}
sessionInfo()
```
