---
title: "scatools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scatools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
We start by loading `scatools` and `ArchR`, as well as the filepath to an example `fragments.bed.gz` file containing fragments from 100 normal mammary cells. This bed file was generated using the `reformatFragmentFiles()` function from the `ArchR` package.

Note this vignette is currently set up with dummy normal data and is purely intended to demonstrate package functionality. There are no CNV changes in the test data.

## Installation

You can install the development version of scatools from [GitHub](https://github.com/) with:

``` r
devtools::install_github("mjz1/scatools")
```

## Example

We start by loading `scatools` and `ArchR`, as well as the filepath to an example `fragments.bed.gz` file containing fragments from 100 normal mammary cells. This bed file was generated using the `reformatFragmentFiles()` function from the `ArchR` package.

Note that all steps in this vignette work with a list of fragment files from multiple samples as well.
```{r setup, eval=TRUE, message=FALSE}
library(scatools)
```


```{r, eval=FALSE}
ncores <- 4 # Adjust accordingly
addArchRThreads(ncores)
addArchRGenome("hg38")

fragment_file <- system.file("extdata", "fragments.bed.gz", package = "scatools")
names(fragment_file) <- "test_sample"
```


Now we process this data using `scatools`. Helper functions help us to create `GenomicRanges` bins, and compute GC content for downstream usage. Here we demonstrate using 10Mb bins.
```{r}
# Generate bins
bins <- get_tiled_bins(bs_genome = BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38, tilewidth = 1e7)
bin_name <- prettyMb(getmode(width(bins)))
```

A convenience wrapper is provided to perform the analysis
```{r, eval=FALSE}
sce <- run_scatools(sample_id = names(fragment_file), 
                    fragments = fragment_file, 
                    outdir = "./example/", 
                    verbose = TRUE, 
                    overwrite = FALSE, 
                    force_arrow = FALSE,
                    ncores = ncores, 
                    bins = bins)
```


```{r, include=FALSE}
# Sideload the SCE object
sce <- get(data(test_sce))
rm(test_sce)
```


Plot the results using the `dittoSeq` package.
```{r}
library(dittoSeq)
library(patchwork)

p1 <- dittoDimPlot(sce, var = "clusters")
p2 <- dittoDimPlot(sce, var = "tumor_cell") + scale_color_manual(values = col_tumor_cells(), name = "Tumor cell")

p1 + p2 & theme(aspect.ratio = 1)
```


We can visualize the results as a heatmap.
```{r cnaheatmap_plot, message=FALSE}
library(ComplexHeatmap)

sce <- sce[,order(sce$tumor_cell, sce$clusters)]

col_clones <- dittoColors()
col_clones <- col_clones[1:length(unique(sce[["clusters"]]))]
names(col_clones) <- levels(factor(sce[["clusters"]]))


left_annot <- ComplexHeatmap::HeatmapAnnotation(
  tumor_cell = sce[["tumor_cell"]],
  cnv_cluster = sce[["clusters"]],
  col = list(tumor_cell = col_tumor_cells(), cnv_cluster = col_clones),
  which = "row"
)


p_ht_cna <- cnaHeatmap(sce, assay_name = "segment_merged_logratios", clust_annot = left_annot, col_fun = logr_col_fun())
p_ht_cna
```


Or plot individual cells
```{r cell_cna_plot}
# Plot the first five cells
plot_cell_cna(sce = sce, cell_id = 1:5, assay_name = "logr_modal")
```
