---
title: "scatools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scatools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

We start by loading `scatools` and `ArchR`, as well as the filepath to an example `fragments.bed.gz` file containing fragments from 100 normal mammary cells. This bed file was generated using the `reformatFragmentFiles()` function from the `ArchR` package.

Note this vignette is currently set up with dummy normal data and is purely intended to demonstrate package functionality. There are no CNV changes in the test data.

## Installation

You can install the development version of scatools from [GitHub](https://github.com/) with:

``` r
devtools::install_github("mjz1/scatools")
```

## Example

We start by loading `scatools`, as well as the filepath to an example `fragments.bed.gz` file containing scATAC fragments from 100 normal mammary cells.

```{r setup, eval=TRUE, message=FALSE, warning=FALSE}
library(scatools)
library(dittoSeq)
library(ComplexHeatmap)
library(patchwork)

ncores <- 4 # Adjust accordingly

fragment_file <- system.file("extdata", "fragments.bed.gz", package = "scatools")
```

Now we process this data using `scatools`. Helper functions help us to create `GenomicRanges` bins, and compute GC content for downstream usage. Here we demonstrate using 10Mb bins.

**Note**: To generate your own `bins` object, a helper function is provided. Please note that running this function requires installation of a `BSgenome` of your choice. Here we use hg38.

```{r, eval=FALSE}
# if (!require("BiocManager", quietly = TRUE))
    # install.packages("BiocManager")

# BiocManager::install("BSgenome.Hsapiens.UCSC.hg38")

bins <- get_tiled_bins(bs_genome = BSgenome.Hsapiens.UCSC.hg38, 
                       tilewidth = 1e7, 
                       respect_chr_arms = TRUE
                       )
```

```{r, include=FALSE}
# Sideload bins
bins <- get(data(bins_10mb))
rm(bins_10mb)
```

We load blacklist regions for our genome. These are used to remove fragments found in low mappability or high signal regions as defined [here](https://github.com/Boyle-Lab/Blacklist).

```{r, eval=TRUE}
# Get blacklist regions
blacklist <- get_blacklist(genome = "hg38")
```

Then we run the pipeline:

```{r, eval=FALSE}
sce <- run_scatools(
  sample_id = "test_sample",
  fragment_file = fragment_file,
  blacklist = blacklist,
  outdir = "./example/",
  verbose = TRUE,
  overwrite = TRUE,
  ncores = ncores,
  bins = bins
)
```

```{r, include=FALSE}
# Sideload the SCE object
sce <- get(data(test_sce))
rm(test_sce)
```

Plot the results using the `dittoSeq` package.

```{r, fig.width = 8, fig.height = 3}
p1 <- dittoDimPlot(sce, var = "clusters")
p2 <- dittoDimPlot(sce, var = "tumor_cell")

p1 + p2 & theme(aspect.ratio = 1)
```

We can visualize the results as a heatmap.

```{r cnaheatmap_plot, message=FALSE, fig.width = 10, fig.height = 5}
sce <- sce[, order(sce$tumor_cell, sce$clusters)]

col_clones <- dittoColors()
col_clones <- col_clones[1:length(unique(sce[["clusters"]]))]
names(col_clones) <- levels(factor(sce[["clusters"]]))


left_annot <- ComplexHeatmap::HeatmapAnnotation(
  tumor_cell = sce[["tumor_cell"]],
  cnv_cluster = sce[["clusters"]],
  col = list(cnv_cluster = col_clones),
  which = "row"
)


p_ht_cna <- cnaHeatmap(sce, assay_name = "segment_merged_logratios", clust_annot = left_annot, col_fun = logr_col_fun())
p_ht_cna
```


Or plot individual cells

```{r cell_cna_plot, fig.width = 9, fig.height = 6, warning=FALSE}
# Plot the first five cells
plot_cell_cna(sce = sce, cell_id = 1:5, assay_name = "logr_modal")
```
